### 2008-09-17 trunk/Papers/rmrs-08/rmrs.tex
iterated over whole paper

\rmrs\ as presented here combines similarities to earlier definitions
of \rmrs\ (e.g., \cite{copestake:2003}) as illustrated in
(\ref{eg:cat}c) and
to dominance constraints
\cite{egg:etal:2001}.  We
$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
$\ARG_\N$ instead of $\ARG_n$.  Furthermore, the
elementary predication $\ep{l}{a}{P(v)}$ is a notational variant of
$\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.
In other words, the syntax defined
here differs in some details, but is more general. The primary
differences to dominance constraints are the distinction of labels and
anchors that will separate the
labelling and the parent-child relations, and
the use of $\qeq$ instead of dominance.  Redefining
\rmrs\ in the
style of dominance constraints will simplify the formal work
below.

\todo{don't split stuff up into tuple parts}

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$ (see Figure~\ref{fig:1}).  So models of the \rmrs\ are
such trees, and we 
can define satisfaction and entailment over these
models in the usual way.  Satisfaction corresponds to 
an \rmrs\ being a partial description of the (unique) fully specific
semantic representation that the model corresponds to, whereas
entailment corresponds to the relation between
a more and a less specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic; in this case, $\tau$ is simply the structure tree of
the predicate logic formula.  We assume that
$\Sigma$ is a ranked signature consisting of predicate logic's
symbols:
a unary constructor $\neg$ and binary constructors $\wedge$,
$\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
x$ for all base-language variables $x$; all the constants and
base-language variables as nullary constructors; and, for each
predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
Other base languages may require a different signature $\Sigma$ and/or
a different mapping between formulas and trees.  We
only assume for convenience that the signature contains a binary
constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
for the set of all constructors in $\Sigma$ that have arity $i$.

@@

The {\em satisfaction} relation between an {\sc rmrs} $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, we assume a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
maps the base variables in the \rmrs\ to nullary constructors
representing variables in the base language.  This function enables us
to map two variables that seem different in the \rmrs\ to the same
base-language variable, introducing e.g.\ semantic dependencies that a
POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
in (\ref{eg:cat-pos}b) to $y$ to specify a variable binding.  Finally,
we assume a relation $\sigma \subseteq \pred 
\times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
a set of constructors from $\Sigma$.  This relation allows us to
underspecify lexical ambiguities; for instance,
$\sigma(\mathrm{\_cat\_n})$ might contain the symbol $\mathrm{\_cat\_n\_1}$
in (\ref{eg:cat}c) along with symbols for other word senses.

@@

Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
This \rmrs\ is satisfied
by the tree in Fig.~\ref{fig:1} (among others) together with some
particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
map the variables $l_1$ and $a_1$ to the root node; $l_{41}$ and
$l_{42}$ to its second child (labeled with ``$\wedge$''); $a_{41}$ to
the first child of \emph{that} node (i.e.\ the node 21, labelled with
``fat'') and $a_{42}$ to the node 22; and so forth.  $g$ will map $x$
and $x_3$ to $x$, $e'$ to $e$, and so on.  And $\sigma$ will map each
\rmrs\ predicate symbol (which represents a word) to the set of its
word senses, e.g.\ \_cat\_n to a set containing \_cat\_n\_1 and
possibly others.  It is then easy to verify that every single atom in
the \rmrs\ is satisfied -- most interestingly, the EPs
$\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``every fat black cat chased some
dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
the restrictor of the universal quantifier.  It is still a model of
(\ref{eg:cat-pos}b)---all the EPs are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula).
The \rmrs\ didn't specify that this material should be
there; but it also didn't specify that it can't be there, and indeed
there is no way in which the \rmrs\ can enforce that no extra material
can be added.

\begin{figure}
  \centering
  \includegraphics[width=6cm]{pic-more-stuff}
  \caption{Another tree which satisfies (\ref{eg:cat-pos}b).}
  \label{fig:fat-black-cat}
\end{figure}

@@

\begin{prop}
  Every RMRS in solved form is satisfiable.
\end{prop}
\begin{proof}[Proof (sketch; see also
  \cite{Duchier00dominanceconstraints})] 
  For each EP, we choose to label the anchor with the constructor $p$
  of sufficiently high arity whose existence we assumed; we determine
  the edges between an anchor and its children from the uniquely
  determined $\ARG$ atoms; plugging labels into holes is
  straightforward because no label is qeq-dominated by more than one
  label; and then we can fill the spaces between the labels and
  anchors with conjunctions.
\end{proof}

@@

\begin{prop}
  For every tuple $(\tau,\alpha,g,\sigma)$ that satisfies some \rmrs\
  $\varphi$, there is a solved form $\varphi'$ of $\varphi$ such that
  $(\tau,\alpha,g,\sigma)$ also satisfies $\varphi'$.
\end{prop}
\begin{proof}
  We construct the substitution $s$ from $\alpha$ and $g$.  Then we
  add all qeqs that are satisfied by $\alpha$ and restrict the $\ARG$
  atoms to those child indices that are actually used in $\tau$.  The
  result is in solved form because $\tau$ is a tree; it is a solved
  form of $\varphi$ by construction.
\end{proof}

@@

\begin{prop}
  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
  renaming of variables.
\end{prop}
\begin{proof}
  Up to renaming of variables, there is only a finite number of
  substitutions on the node and base variables of $\varphi$.  Let $s$
  be such a substitution.  This fixes the set of EPs of any solved
  form of $\varphi$ that is based on $s$ uniquely.  There is only a
  finite set of choices for the subsets $S'$ in condition 2 of
  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
  the set of solved forms of $\varphi$ is finite.
\end{proof}

>>>>

\rmrs\ as presented here combines similarities to earlier
presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
constraints \cite{egg:etal:2001}.  We $\ARG_{\{i\}}$ (with a singleton
set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
Furthermore, the elementary predication $\ep{l}{a}{P(v)}$ is a
notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.  In other
words, the syntax defined here differs in some details, but is more
general. The primary differences to dominance constraints are the
distinction of labels and anchors that will separate the labelling and
the parent-child relations, and the use of $\qeq$ instead of
dominance.  Redefining \rmrs\ in the style of dominance constraints
will simplify the formal work below.

\todo{we don't split stuff up into tuple parts, which earlier rmrs
  did}

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$, such as the one in Fig.~\ref{fig:1}.  So models of the
\rmrs\ are such trees, and we can define satisfaction and entailment
over these models in the usual way.  Satisfaction corresponds to an
\rmrs\ being a partial description of the (unique) fully specific
semantic representation that the model corresponds to, whereas
entailment corresponds to the relation between a more and a less
specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic; in this case, $\tau$ is simply the structure tree of
the predicate logic formula.  We assume that $\Sigma$ is a ranked
signature consisting of predicate logic's symbols: a unary constructor
$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; the
binary constructors $\forall x$ and $\exists x$ for all base-language
variables $x$; all the constants and base-language variables as
nullary constructors; and, for each predicate symbol $P$ of arity $n$,
a constructor $P$ of arity $n$.  Other base languages may require a
different signature $\Sigma$ and/or a different mapping between
formulas and trees.  We only assume for convenience that the signature
contains a binary constructor $\wedge$ to represent conjunction.  We
write $\Sigma_i$ for the set of all constructors in $\Sigma$ that have
arity $i$.  We will follow the typographical convention that
non-logical symbols in $\Sigma$ are written in
$\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
like ``$\sempred{\_cat\_n}$''.

@@

The {\em satisfaction} relation between an \rmrs\ $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, we assume a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
maps the base variables in the \rmrs\ to nullary constructors
representing variables in the base language.  This function enables us
to map two variables that seem different in the \rmrs\ to the same
base-language variable, introducing e.g.\ semantic dependencies that a
POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
in (\ref{ex:cat-pos}) to $y$ to specify a variable binding.  Finally,
we assume a relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$
that maps each \rmrs\ predicate symbol to a set of constructors from
$\Sigma$.  This relation allows us to underspecify lexical
ambiguities; for instance, $\sigma(\sempred{\_cat\_n})$ might contain
the symbol $\sem{\_cat\_n\_1}$ along with symbols for other word
senses.

@@

Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
This \rmrs\ is satisfied by the tree in Fig.~\ref{fig:1} (among
others) together with some particular $\alpha$, $g$, and $\sigma$.
The assignment $\alpha$ will map the variables $l_1$ and $a_1$ to the
root node; $l_{41}$ and $l_{42}$ to its second child (labeled with
``$\wedge$''); $a_{41}$ to the first child of \emph{that} node (i.e.\
the node 21, labelled with ``fat'') and $a_{42}$ to the node 22; and
so forth.  $g$ will map $x$ and $x_3$ to $x$, $e'$ to $e$, and so on.
And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
word) to the set of its word senses, e.g.\ \_cat\_n to a set
containing \_cat\_n\_1 and possibly others.  It is then easy to verify
that every single atom in the \rmrs\ is satisfied -- most
interestingly, the EPs $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``every fat black cat chased
some dog'', differs from the model in Fig.~\ref{fig:1} in the subtree
below the restrictor of the universal quantifier.  It is still a model
of (\ref{ex:cat-pos})---all the EPs are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula).  The \rmrs\
didn't specify that this material should be there; but it also didn't
specify that it can't be there, and indeed there is no way in which
the \rmrs\ can enforce that no extra material can be added.

\begin{figure}
  \centering
  \includegraphics[width=6cm]{pic-more-stuff}
  \caption{Another tree which satisfies (\ref{ex:cat-pos}).}
  \label{fig:fat-black-cat}
\end{figure}

@@

\begin{prop} \label{prop:solved-forms-are-satisfiable}
  Every RMRS in solved form is satisfiable.
\end{prop}
\begin{proof}[Proof (sketch; see also
  \cite{Duchier00dominanceconstraints})] 
  For each EP, we choose to label the anchor with the constructor $p$
  of sufficiently high arity whose existence we assumed; we determine
  the edges between an anchor and its children from the uniquely
  determined $\ARG$ atoms; plugging labels into holes is
  straightforward because no label is qeq-dominated by more than one
  label; and then we can fill the spaces between the labels and
  anchors with conjunctions.
\end{proof}

@@

\begin{prop} \label{prop:models-satisfy-solved-forms}
  For every tuple $(\tau,\alpha,g,\sigma)$ that satisfies some \rmrs\
  $\varphi$, there is a solved form $\varphi'$ of $\varphi$ such that
  $(\tau,\alpha,g,\sigma)$ also satisfies $\varphi'$.
\end{prop}
\begin{proof}
  We construct the substitution $s$ from $\alpha$ and $g$.  Then we
  add all qeqs that are satisfied by $\alpha$ and restrict the $\ARG$
  atoms to those child indices that are actually used in $\tau$.  The
  result is in solved form because $\tau$ is a tree; it is a solved
  form of $\varphi$ by construction.
\end{proof}

@@

\begin{prop}  \label{prop:finite-solved-forms}
  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
  renaming of variables.
\end{prop}
\begin{proof}
  Up to renaming of variables, there is only a finite number of
  substitutions on the node and base variables of $\varphi$.  Let $s$
  be such a substitution.  This fixes the set of EPs of any solved
  form of $\varphi$ that is based on $s$ uniquely.  There is only a
  finite set of choices for the subsets $S'$ in condition 2 of
  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
  the set of solved forms of $\varphi$ is finite.
\end{proof}
